EDA Project by Ahmed
========================================================

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,warning = FALSE, error = FALSE,
                      message = FALSE,cache=TRUE)
```

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
library(ggplot2) # for general plots.
library(dplyr) # for data sets manipulations.
library(gridExtra) # for making a grid of plots.
library(corrplot) # for making a correlation table.
```

```{r echo=FALSE, Load_the_Data}
# pld stands for prosperLoanData
pld <- read.csv('prosperLoanData.csv')
```

# Understanding the data set 
> ** Understanding the data set**:
  This data set contains 113,937 loans with 81 variables on each loan, 
  including loan amount, borrower rate (or interest rate), current loan status,
  borrower income, borrower employment status, borrower credit history, and the 
  latest payment information.
  
  [This](https://docs.google.com/spreadsheets/d/1gDyi_L4UvIrLTEC6Wri5nbaMmkGmLQBk-Yx3z0XDEtI/edit?usp=sharing) is the Variables dictionary which contains helpful info about each summary.
some interesting variables that we could look at: 
- IncomeRange
- BorrowerRate
- IsBorrowerHomeOwner
- the BorrowerAPR
- the Employment status
- LoanStatus
- LoanDuration
- LenderYield
- Grade of Borrowers.
- Credit Score

### the structure  of the data set.
```{r echo=FALSE}
View(head(pld))
str(pld)
```

# Univariate Analysis

### the income Range
```{r echo=FALSE}
# removed the Not Displayed values.
ggplot(aes(x = IncomeRange), data = subset(pld, pld$IncomeRange != 'Not displayed'))+
  geom_bar( fill='yellow',color = 'black',size =1)
  summary(pld$IncomeRange)
```

> most of the Borrowers are in the $25,000-49,999 segment.

### the number of home owners.
```{r echo=FALSE}
ggplot(aes(x = IsBorrowerHomeowner, ),data = pld)+
  geom_bar(fill = c('red','green'))
summary(pld$IsBorrowerHomeowner)
```

> the number of those who own homes and those who don't is very close.

### the Borrwoer rate interest
```{r echo=FALSE}
ggplot(aes(x=BorrowerRate),data=pld)+
  geom_histogram(binwidth = 0.02,color='black',fill='white')
print('the Mean is : ')
mean(pld$BorrowerRate)
```

> this histogram displays the Borrower interest rate distribution of the Borrowers interest Rate, we can see that the average loan interest is 19.28 %  and here's the summary of the distribution including the five numbers summary.

```{r echo=FALSE}
summary(pld$BorrowerRate)
```

### employment status
```{r echo=FALSE}
pld$EmploymentStatus <- factor(pld$EmploymentStatus,levels = 
ordered(c('Not employed',' Retired','Part-time',
          'Self-employed','Full-time','Employed')))
ggplot(aes(x=EmploymentStatus),data= subset(pld,!pld$EmploymentStatus == ''))+
  geom_bar(color='green',fill='brown',size=0.8)+
  xlab('Employment Status')
summary(pld$EmploymentStatus)
```

> its obvious now that most of the Borrowers are employed.

```{r echo=FALSE}

ggplot(aes(x=BorrowerAPR),data=pld)+
geom_bar(binwidth = 0.02,color='red',fill='black')
ggplot(aes(x=BorrowerAPR),data=pld)+
geom_bar(binwidth = 0.02,color='red',fill='black')

print("The APR summary statistics is :")
summary(pld$BorrowerAPR)
```

> The loan APR ranges from less than 5 % up to about 51 %.
Most borrowers seems to have a APR from 10 % upto 30 % with average equals to 21.9 %.

### Loan Status
```{r echo=FALSE}
pld$LoanStatus <- factor(pld$LoanStatus, levels = 
ordered(c("Cancelled","Chargedoff", "Completed", "Current", "Defaulted","FinalPaymentInProgress","PastDue.")))

ggplot(aes(x=LoanStatus),data = subset(pld,!is.na(LoanStatus)))+
  geom_bar(fill='gray',color='red')+
  xlab('Loan Status')
# displlaying the percentage of each category using dplyr.
pld_status_table=table(pld$LoanStatus)
x=data.frame(pld_status_table)
names(x)=c('LoanStatus', 'Percent')
category=data.frame(table(pld$LoanStatus))
x$Percent= round(category$Freq/sum(category$Freq)*100 , 2)
print(x)
```

> most of the borrowers are still currently taking loans with a percentage of 50.57%, 34.03% has already completed, and 10.72% are charged off!

### Loan duration
```{r echo=FALSE}
ggplot(aes(x=as.factor(Term)),data=subset(pld,!is.na(Term)))+
  geom_bar(fill='pink')+xlab("Term (months)")
```

> 36 months is most common Term.


### ProsperRating Grade of the borrowers 
```{r}

pld$ProsperRating..Alpha.=factor(pld$ProsperRating..Alpha.,levels =
      ordered(ordered(c('AA','A','B','C', 'D','E','HR',""))))
ggplot(aes(x=ProsperRating..Alpha.),data=
subset(pld,ProsperRating..Alpha. %in% ordered(c('AA','A','B','C', 'D','E','HR'))))+ 
  geom_bar(color='black',fill='dark green',size=1.25)+
  xlab('Borrower Grade')
```

> the Grade is normally distributed with a mode equals to C.

### Lender Yield.
```{r}
ggplot(aes (x = LenderYield), data = pld) +
  geom_histogram(color='red',fill = 'white')
summary(pld$LenderYield)                                        
```

> Lender yield is equal to the interest rate on the loan less the servicing fee
and it's average equals 18.27% .

### Borrower APR.

```{r echo = FALSE}

ggplot(aes(x=BorrowerAPR),data=pld)+
geom_bar(binwidth = 0.02,color='red',fill='maroon') +
  labs(x="Borrower APR",y="Count",title="Borrower APR distribution")
summary(pld$BorrowerAPR)
```
this plot shows the distribution of the Borrower APR Variable with an average
equals to 21.88%.

# Bivariate Analysis
> some questions in my mind right now..
- is there a relationship between being a home owner and the loan status? 
probably the home owners complete their loans more quickly!?

### home owner by current loan status

```{r echo=FALSE}

by(pld$LoanStatus ,pld$IsBorrowerHomeowner,summary)

p1 <- ggplot(aes(x=IsBorrowerHomeowner),data=
    subset(pld, (LoanStatus =='Current' )))+
  geom_bar(fill='green2')+xlab('currently loaners and Has a Home')
p2 <- ggplot(aes(x=IsBorrowerHomeowner),data =
subset(pld, (LoanStatus=='Completed' )))+
geom_bar(fill = 'green2')+xlab('Completed the loan and Has a Home')

p3 <- ggplot(aes(x=IsBorrowerHomeowner),data=
subset(pld, !(LoanStatus=='Chargedoff' | LoanStatus == 'Defaulted')))+
  geom_bar(fill='blue')+xlab('chargedoff or Defaulted and Has a Home')

grid.arrange(p1,p2,p3, ncol = 2)
```

a plot shows the relation between the Home owners and the Current status of 
loans.


### Estimated Effective Yield, EstimatedReturn and Estimated Loss 

> - is there a relation betweem the Effective Yield and the Return or the Loss?
```{r}
p1 <- ggplot(data=pld, aes(x=EstimatedEffectiveYield,y=EstimatedReturn))+
  geom_point(color='dark green',alpha=0.1)+
  xlab('Estimated Effective Yield')+
  ylab('Estimated Return')

p2 <- ggplot(data=pld, aes(x=EstimatedLoss,y=EstimatedReturn))+
  geom_point(color='red',alpha=0.1)+
  xlab('Estimated Loss')+
  ylab('')
grid.arrange(p1,p2,ncol=2)
```

There's a positive relation between the the Estimated Return and the Estimated 
EffectiveYield whereas there's a negative relation between the Return and the
Estimated 
Loss.

### Monthly payments with loan amount

> - probably the monhly payments has a relation with the loans amout, let's make
sure!?

```{r echo=FALSE}
ggplot(aes(x=LoanOriginalAmount,y=MonthlyLoanPayment),data=pld)+
  geom_point(color='Green',alpha=1/2)
```

obviously the Higher Loans are correlated with the monthly Payments.


### Estimated loss and Prosper rating
> - bad ratings probably have a relation with the Estimated loss?

```{r echo=FALSE}
pld$ProsperRating..Alpha. <- factor(pld$ProsperRating..Alpha., levels = 
  ordered(rev(c('AA','A', 'B','C','D','E','HR'))))
ggplot(aes(x=ProsperRating..Alpha., y=EstimatedLoss),
       data=subset(pld,!is.na(ProsperRating..Alpha.)))+
geom_boxplot(aes(fill=ProsperRating..Alpha.))+
  guides(fill=guide_legend(title = 'Rating'))
```

> Yep it's a negative correlation, the bad ratings are associated with High
Estimated Loss.


### Prosper rating and credit score 
> is there's a relation between the Prosper Ratin and the Credit score?

```{r echo=FALSE}
pld$ProsperRating..Alpha. <- factor(pld$ProsperRating..Alpha., levels = 
                              ordered(rev(c('AA','A', 'B','C','D','E','HR'))))
p1 <- ggplot(aes(x=ProsperRating..Alpha., y=CreditScoreRangeLower),
             data=subset(pld,!is.na(ProsperRating..Alpha.)))+
  geom_boxplot(aes(fill=ProsperRating..Alpha.))+
  guides(fill=guide_legend(title = 'Rating'))+
  xlab('Prosper Rating')+
  ylab('Credi Score Range Lower')

p2 <- ggplot(aes(x=ProsperRating..Alpha., y=CreditScoreRangeUpper),
             data=subset(pld,!is.na(ProsperRating..Alpha.)))+
  geom_boxplot(aes(fill=ProsperRating..Alpha.))+
  guides(fill=guide_legend(title = 'Rating'))+
  xlab('Prosper Rating')+
  ylab('Credi Score Range Upper')
grid.arrange(p1,p2,ncol=1)
```

Credit scores have an impact on the proposer rating, as the credit score 
increases the Rating increases too which is too obvious.


### APR and Prosper rating
> - now let's look at the APR, is there a relation between it and the Rating?

```{r echo=FALSE}
ggplot(aes(ProsperRating..Alpha.,y=BorrowerAPR),data = 
         subset(pld,!is.na(ProsperRating..Alpha.)))+
geom_boxplot(aes(fill=ProsperRating..Alpha.))+
  xlab('Prosper Rating')
```  

borrowers with good rating have lower APRs and those with 
bad ratings have high APRs.



### APR and interest rate
```{r echo=FALSE}
ggplot(aes(x=BorrowerAPR,y=BorrowerRate),data=pld)+geom_point(fill=I('#F79420'),
                                                             color=I('black'),shape=21,alpha=0.5)+geom_smooth(method='lm',color='dark green')
```

```{r echo=FALSE, Bivariate_Plots}
print("The summary statistics of borrower APR is ")
summary(pld$BorrowerAPR)
print("The summary statistics of borrower rate is ")
summary(pld$BorrowerRate)
```

The borrower APR and rate are linearly related. The average APR and rate 
are ~ 22% and 19 % respectively.
### Loan Amount vs loan term
```{r echo=FALSE}
ggplot(aes(x=Term,y=LoanOriginalAmount),data = pld)+
  geom_boxplot(aes(fill=factor(Term)))+
  stat_summary(fun.y=mean, geom="point", shape=8)
```

```{r echo=FALSE}
summary(pld$LoanOriginalAmount)
```

Bigger Terms are associated with the big Amounts of Loans. 
Using the summary stats, we can see the average loan amount is $8337 and the 
loan amount ranges between $1000 and $35000.



# MULTIVARIATE ANALYSIS

### Monthly payement by total loan and Loan Term

```{r echo=FALSE}
ggplot(aes(x=LoanOriginalAmount,y=MonthlyLoanPayment),data=pld)+
  geom_point(aes(color=factor(Term)),alpha=1/2)

```

```{r}
summary(pld$MonthlyLoanPayment)
```

we have seen that the monthly payment increases as the amount of loan gets 
bigger.For the same loan amount the monthly payment amount is higher for shorter
terms. From the summary statistics we can see that 
the average monthly payment is almost $272.


### Borrower classification by loan range and loan term?

```{r}
pld$LoanRange <- cut(as.numeric(as.character(pld$LoanOriginalAmount)),breaks =
                      seq(0, 35000, by = 5000))
pld$LoanRange <- factor(pld$LoanRange,levels=c('(0,5e+03]',   '(5e+03,1e+04]',
'(1e+04,1.5e+04]', '(1.5e+04,2e+04]', '(2e+04,2.5e+04]', '(2.5e+04,3e+04]', 
'(3e+04,3.5e+04]' ),ordered=TRUE)
k=table(pld$LoanRange)
kk=data.frame(k)
colnames(kk)=c('Loan_range','Number_of_borrowers')
print(kk)

ggplot(aes(x=LoanRange),data =pld)+geom_bar(aes(fill=factor(Term)))+
ylab("Number of borrowers ")+xlab('Loan range')+
  theme(axis.text.x = element_text(angle = 90, hjust = 1))#+geom_freqpoly()
```

It is clear to see that most borrowers have loans in the range from 1-5000 for 
aperiod of 36 months. Generallly most borrowers borrowed for 36 and 60 months 
and there are no borrowers borrowed more than 15000 for a period of 12 months.

### Distribution of  borrowers by loan range and employment status? 

```{r echo=FALSE}
ggplot(aes(x=LoanRange),data =subset(pld,!is.na(EmploymentStatus)))+
  geom_bar(aes(fill=factor(EmploymentStatus)))+
  ylab("Number of borrowers ")+xlab('Loan range')+
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

most prosper loan amounts are in the range of 0,5000 followed by 5000 - 10,000 
and most borrowers are employed. 

### Effective yield with rating and term

```{r}
ggplot(aes(ProsperRating..Alpha.,y=EstimatedLoss),data = subset(pld,!is.na(ProsperRating..Alpha.)))+geom_boxplot(aes(fill=factor(Term)))+
xlab('Prosper rating')+ylab('Estimated loss')
summary(pld$EstimatedLoss)
```

The Estimated Loss Decreases as the Rating increases nd the Term is almost the 
same for all groups.
### APR with term and rating

```{r}
ggplot(aes(ProsperRating..Alpha.,y=BorrowerAPR),data = subset(pld,!is.na(ProsperRating..Alpha.)))+geom_boxplot(aes(fill=factor(Term)))
``` 

as the Rating Increases the APR decreases, and The 12 Months term members is
kinda Decreasing until reaching a very small Ratio at the AA Rate.


### Correlation Table between different  variables

```{r  echo=FALSE}
new_pld <- pld[, c("EstimatedEffectiveYield","EstimatedLoss","EstimatedReturn",
"ProsperRating..numeric.", "EmploymentStatusDuration","CreditScoreRangeLower")]
m <- cor(na.omit(new_pld))#[,1:8])
corrplot(m,method = 'number')
```

a correlation table of some variables.


# FINAL PLOTS 

### Plot 1

```{r echo = FALSE}
ggplot(aes(x=BorrowerAPR),data=pld)+
geom_bar(binwidth = 0.02,color='red',fill='maroon') +
geom_vline(xintercept = 0.21880, color='black',linetype='dotted',size=1.2)+
geom_text(aes(x=0.24, label="Average APR", y=10000), colour="black",
          angle=90,text=element_text(size=11)) +
  labs(x="Borrower APR",y="Count",title="Borrower APR distribution")+
  theme(plot.title = element_text(hjust = 0.5))
```

The loan APR ranges from less than 5 % up to about 51 %. However, there are only 
very few borrowers with such high and lower APR values. Most borrowers seems to
have a APR from 10 % to 30 %. The median APR value is ~21 %. 

### plot 2
```{r echo=FALSE}

by(pld$LoanStatus ,pld$IsBorrowerHomeowner,summary)

p1 <- ggplot(aes(x=IsBorrowerHomeowner),data=
    subset(pld, (LoanStatus =='Current' )))+
  geom_bar(fill='green2')+xlab('currently loaners and Has a Home')
p2 <- ggplot(aes(x=IsBorrowerHomeowner),data =
subset(pld, (LoanStatus=='Completed' )))+
geom_bar(fill = 'green2')+xlab('Completed the loan and Has a Home')

p3 <- ggplot(aes(x=IsBorrowerHomeowner), data=
subset(pld, !(LoanStatus=='Chargedoff' | LoanStatus == 'Defaulted'))) +
  geom_bar(fill='blue')+xlab('chargedoff or Defaulted and Has a Home')

grid.arrange(p1,p2,p3, ncol = 2)
```

> I was thinking that most of those who completed the loan should be with home
but this has changed my opinion!, most of those who completed already are with
no homes however those who have homes most of them still currently loaners,
changed off or Defaulted.

### plot 3

```{r echo=FALSE}
ggplot(aes(x=factor(Term),y=LoanOriginalAmount),data = pld)+
  geom_boxplot(aes(fill=factor(Term)))+
  stat_summary(fun.y=mean, geom="point", shape=8)+
  ylab('Loan Orignal Amount (USD)')
```

```{r echo=FALSE}
summary(pld$LoanOriginalAmount)
```

Bigger Terms are associated with the big Amounts of Loans. 
Using the summary stats, we can see the average loan amount is $8337 and the 
loan amount ranges between $1000 and $35000.


#REFLECTION

The Prosper loan data set has 113,937 rows and 81 features. I explored some of 
the 81 variables and tried to find interesting relationships among the  
variables. My focus of investigation was mostly on the profile of the borrowers 
such as income range, employment, credit score etc.. and the features of the 
loan such as APR, estimated loss, loan amount with Term etc..,. I used ggplot to
check how these variable distributed and related.
On the process of exploration, 
I found many missing values in the data and I handeled the problem using R
data cleaning tools such as from the subset function.
 
From the exploration of the data, I found the APR varies from less than 1 % upto
a little higher than 50 %. 
The APR is related to the credit score of the borrowers. It is foud that 
borrowers with high credit scores have good prosper rating and lower APR. It is 
also found that most borrowers have loans in the range from 1-5000 for aperiod 
of 36 months.
Finally I plotted the correlation table of many variables and seen many 
interesting correlations.

# REFERENCES
https://docs.google.com/document/d/e/2PACX-1vRmVtjQrgEPfE3VoiOrdeZ7vLPO_p3KRdb_o-z6E_YJ65tDOiXkwsDpLFKI3lUxbD6UlYtQHXvwiZKx/pub?embedded=true
https://s3.amazonaws.com/content.udacity-data.com/courses/ud651/
diamondsExample_2016-05.html
http://stackoverflow.com/